---
title: "Getting Started with structr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with structr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(structr)
```

## Introduction

Have you ever had trouble with data-type consistency when interacting with
external systems like a web API or a database? It is not uncommon for JSON
parsers in R to fail to parse and serialize data to the specification of the
external system.

For example, let's say an API requires a field in a JSON object to be a single
scalar integer value. However, you get constant `400 Bad Request` because
packages like `jsonlite` serialize the integer as a vector of length 1. If the
API requires a nested structure unbox values becomes very impractical.

The `structr` package provides a solution to this problem by allowing you to
define a structure for your data and then serialize and deserialize it to and
from JSON. (With the added benefit of being blazingly fast.)

## Defining a structure

Defining a structure with `structr` involves combining different data-types to build
complex structures. These are combined through the `s_*` functions like
`?s_map`, `?s_vector`, `?s_string` and others.

Once the desired structure is reached it has to be compiled using the
`?build_structure` function. The compiled version is what will be used to parse
or serialize the structures we want.

```{r}
structure_basic <- build_structure(
  s_map(
    x = s_integer(),
    y = s_double(),
    z = s_optional(s_string())
  )
)
print(structure_basic)
```

## Parsing JSON

We can use a structure like the one defined above to parse a JSON exactly to the specification.

```{r}
json_str <- '{"x": 1, "y": 99.9, "z": "value"}'

parse_json(json_str, structure_basic)
```

You may have noticed that `z` is a combination of `?s_optional` and `?s_string`. This is because
`structr` is non-null by default. Every field is required be default, meaning it cannot be `NULL` or `NA`.

If we try parse a JSON without the field `x` we will get an error.

```{r}
json_str <- '{"y": 99.9, "z": "value"}'
try(parse_json(json_str, structure_basic))

# Using 'null' for a non-optional field also causes an error
json_str_null <- '{"x": null, "y": 99.9, "z": "value"}'
try(parse_json(json_str_null, structure_basic))
```

However, if we try to parse a JSON missing the *optional* field `z`, it will not fail. `s_optional` means the *value* associated with the key can be `null` or missing from the JSON object.

```{r}
json_str_missing_z <- '{"x": 1, "y": 99.9}'
parse_json(json_str_missing_z, structure_basic)

json_str_null_z <- '{"x": 1, "y": 99.9, "z": null}'
parse_json(json_str_null_z, structure_basic)
```

We will get errors if the types are incorrect. If we pass a string to x or y we will
get errors.

```{r}
json_str_bad_types <- '{"x": "1", "y": "99.9", "z": "value"}'
try(parse_json(json_str_bad_types, structure_basic))
```

## Serializing JSON

We can use this same structure to convert R objects into the corresponding JSON.
`serialize_json` ensures that the R object conforms to the structure *before*
creating the JSON string.

```{r}
r_list <- list(x = 1L, y = 99.9, z = "value")
serialize_json(r_list, structure_basic)

r_list_null_z <- list(x = 2L, y = 100.1)
serialize_json(r_list_null_z, structure_basic)
```

If the R object doesn't match the structure (e.g., wrong type, missing required field), `serialize_json` will raise an error.

```{r}
# Wrong type for 'x'
r_list_bad_type <- list(x = "1", y = 99.9, z = "value")
try(serialize_json(r_list_bad_type, structure_basic))

# Missing required field 'y'
r_list_missing_field <- list(x = 1L, z = "value")
try(serialize_json(r_list_missing_field, structure_basic))

# Providing NA for a non-optional field 'y'
r_list_na_field <- list(x = 1L, y = NA_real_, z = "value")
try(serialize_json(r_list_na_field, structure_basic))
```


## More Complex Structures (Parsing)

Real-world JSON, especially from APIs, often involves nested objects, arrays of objects, optional fields, and specific data types like dates. `structr` is designed to handle these complexities robustly during parsing.

Let's imagine we're interacting with an e-commerce API that returns order details. A typical JSON response might look like this:

```json
{
  "order_id": "ORD-12345",
  "order_date": "2024-07-29",
  "status": "processing",
  "customer": {
    "customer_id": 987,
    "name": "Jane Doe",
    "email": "jane.doe@example.com"
  },
  "items": [
    {
      "sku": "ITEM001",
      "product_name": "Wireless Mouse",
      "quantity": 1,
      "unit_price": 25.99
    },
    {
      "sku": "ITEM002",
      "product_name": "Keyboard",
      "quantity": 1,
      "unit_price": 75.50
    }
  ],
  "shipping_address": {
    "street": "123 Example St",
    "city": "Metropolis",
    "zip_code": "10001",
    "country": "USA"
  },
  "notes": "Gift wrap requested."
}
```

We can define a `structr` schema to strictly match this structure:

```{r complex_schema_parse}
# Define the schema for a single order item
item_schema_parse <- s_map(
  sku = s_string(),
  product_name = s_string(),
  quantity = s_integer(),
  unit_price = s_double()
)

# Define the schema for the customer
customer_schema_parse <- s_map(
  customer_id = s_integer(),
  name = s_string(),
  email = s_optional(s_string()) # Email might be null
)

# Define the schema for the shipping address
address_schema_parse <- s_map(
  street = s_string(),
  city = s_string(),
  zip_code = s_string(),
  country = s_string()
)

# Combine into the main order schema
order_schema_parse <- s_map(
  order_id = s_string(),
  order_date = s_date(format = "%Y-%m-%d"), # Expecting specific date format
  status = s_string(),
  customer = customer_schema_parse,              # Nest the customer schema
  items = s_vector(item_schema_parse),           # Expect a vector of items matching item_schema
  shipping_address = address_schema_parse,       # Nest the address schema
  notes = s_optional(s_string())           # Notes might be null
)

# Build the final structure
compiled_order_structure_parse <- build_structure(order_schema_parse)

# Print the structure (optional, for inspection)
# print(compiled_order_structure_parse)
```

Now, let's define the JSON string in R and parse it using our compiled structure.

```{r parse_complex_json}
json_order_1 <- '{
  "order_id": "ORD-12345",
  "order_date": "2024-07-29",
  "status": "processing",
  "customer": {
    "customer_id": 987,
    "name": "Jane Doe",
    "email": "jane.doe@example.com"
  },
  "items": [
    {
      "sku": "ITEM001",
      "product_name": "Wireless Mouse",
      "quantity": 1,
      "unit_price": 25.99
    },
    {
      "sku": "ITEM002",
      "product_name": "Keyboard",
      "quantity": 1,
      "unit_price": 75.50
    }
  ],
  "shipping_address": {
    "street": "123 Example St",
    "city": "Metropolis",
    "zip_code": "10001",
    "country": "USA"
  },
  "notes": "Gift wrap requested."
}'

# Parse the JSON string
parsed_order_1 <- parse_json(json_order_1, compiled_order_structure_parse)

# Inspect the parsed R object structure and types
str(parsed_order_1, max.level = 2)

# Access specific parts
print(paste("Order Status:", parsed_order_1$status))
print(paste("Number of items:", length(parsed_order_1$items)))
print(paste("First item SKU:", parsed_order_1$items[[1]]$sku))
# Notice the order_date is now an R Date object
print(paste("Order Date:", parsed_order_1$order_date))
class(parsed_order_1$order_date)
```

Let's try another example where the optional fields (`email` and `notes`) are `null`:

```{r parse_complex_json_nulls}
json_order_2 <- '{
  "order_id": "ORD-67890",
  "order_date": "2024-07-30",
  "status": "shipped",
  "customer": {
    "customer_id": 654,
    "name": "John Smith",
    "email": null
  },
  "items": [
    {
      "sku": "ITEM003",
      "product_name": "USB Hub",
      "quantity": 2,
      "unit_price": 12.00
    }
  ],
  "shipping_address": {
    "street": "456 Test Ave",
    "city": "Gotham",
    "zip_code": "20002",
    "country": "USA"
  },
  "notes": null
}'

# Parse the second JSON
parsed_order_2 <- parse_json(json_order_2, compiled_order_structure_parse)

# Check the optional fields (which are NULL in R)
print(paste("Customer Email:", parsed_order_2$customer$email))
is.null(parsed_order_2$customer$email)
print(paste("Order Notes:", parsed_order_2$notes))
is.null(parsed_order_2$notes)
```

If the incoming JSON deviates from this structure (e.g., wrong type for `quantity`, missing `status`, unexpected extra field, incorrect date format), `parse_json` will raise a specific error, ensuring data integrity when working with external systems.

This example demonstrates how `s_map`, `s_vector`, `s_optional`, `s_date`, and atomic types (`s_string`, `s_integer`, `s_double`) can be combined to model and validate complex, real-world JSON data during parsing.


## More Complex Structures (Serialization)

Just as `structr` helps enforce structure when parsing JSON, it's equally valuable when *creating* JSON to send to external systems, especially those with strict API requirements. This ensures your R data is correctly formatted before transmission.

Let's imagine we need to send event log data to an API. This API is very particular about the format:

*   It expects specific fields: `eventId`, `eventDate`, `eventType`, `userId`, `details`.
*   `eventDate` *must* be in `YYYYMMDD` format.
*   `userId` *must* be an integer.
*   `details` is a nested object containing `source` (string) and `metadata` (optional map).
*   `metadata` itself can contain `key` (string) and `value` (double), both optional within the metadata map, but if `metadata` is present, it must be an object.

We can define a `structr` schema to meet these strict requirements for serialization:

```{r complex_schema_serialize}
# Schema for the optional metadata map
metadata_schema_serialize <- s_map(
  key = s_optional(s_string()),
  value = s_optional(s_double())
)

# Schema for the details object
details_schema_serialize <- s_map(
  source = s_string(),
  metadata = s_optional(metadata_schema_serialize) # The whole metadata map is optional
)

# Main schema for the log entry
log_entry_schema_serialize <- s_map(
  eventId = s_string(),
  eventDate = s_date(format = "%Y%m%d"), # Strict date format required by API
  eventType = s_string(),
  userId = s_integer(),
  details = details_schema_serialize      # Nest the details schema
)

# Build the final structure for serialization
compiled_log_entry_structure_serialize <- build_structure(log_entry_schema_serialize)

print(compiled_log_entry_structure_serialize)
```

Now, let's create some R data that we want to send to this API. It's crucial that the R data types match the schema (e.g., use `as.Date()` for `eventDate`, `integer` for `userId`).

```{r prepare_complex_r_data}
# Example 1: Log entry with full details including metadata
log_entry_data_1 <- list(
  eventId = "evt-aaa-001",
  eventDate = as.Date("2024-08-15"),
  eventType = "user_login",
  userId = 12345L,
  details = list(
    source = "web_frontend",
    metadata = list(
      key = "session_duration",
      value = 350.5
    )
  )
)

# Example 2: Log entry where metadata is missing
log_entry_data_2 <- list(
  eventId = "evt-bbb-002",
  eventDate = as.Date("2024-08-16"),
  eventType = "page_view",
  userId = 98765L,
  details = list(
    source = "mobile_app"
  )
)

# Example 3: Log entry where metadata is present, but its inner fields are NULL
log_entry_data_3 <- list(
  eventId = "evt-ccc-003",
  eventDate = as.Date("2024-08-17"),
  eventType = "item_click",
  userId = 12345L,
  details = list(
    source = "web_frontend",
    metadata = list(
      key = NULL,
      value = NULL
    )
  )
)

str(log_entry_data_1) # Check R types
```

Let's serialize these R lists using our strict structure. `serialize_json` will validate the R list against the schema before producing the JSON.

```{r serialize_complex_json}
# Serialize Example 1
serialize_json(log_entry_data_1, compiled_log_entry_structure_serialize, pretty = TRUE)

# Serialize Example 2
serialize_json(log_entry_data_2, compiled_log_entry_structure_serialize, pretty = TRUE)

# Serialize Example 3
serialize_json(log_entry_data_3, compiled_log_entry_structure_serialize, pretty = TRUE)
```

Now, let's see what happens if our R data *doesn't* conform to the structure expected by the API schema.

```{r serialize_complex_errors}
# Error Case 1: Incorrect date type (character instead of Date)
bad_data_date_type <- log_entry_data_1
bad_data_date_type$eventDate <- "2024-08-15" # Should be as.Date("2024-08-15")
try(serialize_json(bad_data_date_type, compiled_log_entry_structure_serialize))

# Error Case 2: Incorrect type for userId (numeric instead of integer)
bad_data_user_type <- log_entry_data_1
bad_data_user_type$userId <- 12345.0 # Should be 12345L
try(serialize_json(bad_data_user_type, compiled_log_entry_structure_serialize))

# Error Case 3: Missing a required field ('eventType') in the R list
bad_data_missing_field <- log_entry_data_1
bad_data_missing_field$eventType <- NULL # Remove the field
try(serialize_json(bad_data_missing_field, compiled_log_entry_structure_serialize))

# Error Case 4: Providing NA for a required, non-optional field ('source')
bad_data_na_field <- log_entry_data_1
bad_data_na_field$details$source <- NA_character_
try(serialize_json(bad_data_na_field, compiled_log_entry_structure_serialize))
```

These examples show how `serialize_json` uses the `structr` schema to enforce type correctness and structure *before* generating the JSON. This is critical for reliable communication with APIs that demand specific formats, preventing malformed requests and runtime errors on the server side. The use of `s_date` with a specific `format` ensures dates are serialized exactly as required.